---
title: "Exploration of NOAA Storm Database"
output:
  html_document: default
date: "4 May 2017"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## SYNOPSIS
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.  
This report explore U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database and address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
2. Across the United States, which types of events have the greatest economic consequences?.    

## DATA PROCESSING
The data come in the form of a comma-separated-value file.  
Data will be downloaded from: Storm Data [47Mb].  

There is also some documentation of the database available. Here you will find how some  of the variables are constructed/defined.  

- National Weather Service Storm Data Documentation  
- National Climatic Data Center Storm Events FAQ  

From url: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2, the raw data will be downloaded as StormData.csv.bz2

### Loading Data
```{r echo=TRUE}
setwd("D:\\Data Scientist\\Module 5\\Course Project 2")

if(!file.exists("StormData.csv.bz2")) {
  Original_Data_URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(Original_Data_URL, destfile="StormData.csv.bz2")
}
```

### Reading Data & Filter Useful Fields
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)

Stormdata <- read.csv("StormData.csv.bz2", stringsAsFactors=F)

#check number's of row and variable
dim(Stormdata)

# Filter Raw Data: EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP

subset.storm <- Stormdata %>%
  select(STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```


```{r echo=TRUE}
#display 6 row form subset.storm
head(subset.storm)
```

## RESULT

###1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
There are two measurements in the dataset can reflect the degree of harmfulness of a 
type of event with respect to population health: fatalities and injuries.   
Thus, sum them up over types of events to find out the most harmful type of event.  

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Fatalities Category
fatalData <- aggregate(FATALITIES ~ EVTYPE, data = subset.storm, FUN = sum)
table(fatalData$FATALITIES)

# Injuries Category
InjureData <- aggregate(INJURIES ~ EVTYPE, data = subset.storm, FUN = sum)
table(InjureData$INJURIES)

# From two table above, plot the top 5 harmful event for injured and fatal category with reference to population health
library(ggplot2)

PlotFatal <- fatalData[order(fatalData$FATALITIES, decreasing = T), ]
PlotInjured <- InjureData[order(InjureData$INJURIES, decreasing = T), ]

# From two table above, plot the top 5 harmful event for injured and fatal category with reference to population health
# Plot top 5 fatalities per event type
ggplot(PlotFatal[1:5, ], aes(EVTYPE, FATALITIES)) + geom_bar(stat = "identity") + 
  ylab("Number Of Fatalities") + xlab("Event") + ggtitle("Numbers Of Fatalities Per Events Across the U.S")

# Plot top 5 Injured per event type
ggplot(PlotInjured[1:5, ], aes(EVTYPE, INJURIES)) + geom_bar(stat = "identity") + 
  ylab("Number Of Injured") + xlab("Event") + ggtitle("Numbers Of Injured Per Events Across the U.S")

```

Tornado  is the most harmful event type as shown in the above two bar charts, which has led to 5633 deaths and 91346 injuries from year 1950 to November 2011.  

###2. Across the United States, which types of events have the greatest economic consequences?  
   
As we can see, there are two damages PROPDMG and CROPDMG. 
Since the damage costs are reported in two separate columns, a damage and damage exponent   column, create a separate columns in the dataframe to assign the PROPDMGEXP and CROPDMGEXP to the correct value.  
However, it does not account for the characters like "-", "+" or "?" and the numbers like "1", "2", etc.   
   
```{r echo=TRUE}
subset.storm$PROPDMG[subset.storm$PROPDMGEXP %in% "B"] <- subset.storm$PROPDMG[subset.storm$PROPDMGEXP %in% "B"] * 1000
subset.storm$PROPDMG[subset.storm$PROPDMGEXP %in% c("M", "m")] <- subset.storm$PROPDMG[subset.storm$PROPDMGEXP %in% c("M", "m")] * 1
subset.storm$PROPDMG[subset.storm$PROPDMGEXP %in% c("K")] <- subset.storm$PROPDMG[subset.storm$PROPDMGEXP %in% c("K")] * 0.001
subset.storm$PROPDMG[subset.storm$PROPDMGEXP %in% c("H", "h")] <- subset.storm$PROPDMG[subset.storm$PROPDMGEXP %in% c("H", "h")] * 1e-04
subset.storm$PROPDMG[!(subset.storm$PROPDMGEXP %in% c("B", "M", "m", "K", "H", "h"))] <- subset.storm$PROPDMG[!(subset.storm$PROPDMGEXP %in% c("B", "M", "m", "K", "H", "h"))] * 1e-06

subset.storm$CROPDMG[subset.storm$CROPDMGEXP %in% "B"] <- subset.storm$CROPDMG[subset.storm$CROPDMGEXP %in% "B"] * 1000
subset.storm$CROPDMG[subset.storm$CROPDMGEXP %in% c("M", "m")] <- subset.storm$CROPDMG[subset.storm$CROPDMGEXP %in% c("M", "m")] * 1
subset.storm$CROPDMG[subset.storm$CROPDMGEXP %in% c("K", "k")] <- subset.storm$CROPDMG[subset.storm$CROPDMGEXP %in% c("K", "k")] * 0.001
subset.storm$CROPDMG[!(subset.storm$CROPDMGEXP %in% c("B", "M", "m", "K", "k"))] <- subset.storm$CROPDMG[!(subset.storm$CROPDMGEXP %in% c("B", "M", "m", "K", "k"))] * 1e-06
```
Calculated the total damage by adding all property damages and crop damage for the events.  
Then, visualize the top five events.  

```{r echo=TRUE}
EcoConsDmg <- subset.storm$PROPDMG + subset.storm$CROPDMG
EcoCons <- aggregate(EcoConsDmg ~ subset.storm$EVTYPE, FUN = sum)
PlotEcoCons <- EcoCons[order(EcoCons$EcoConsDmg, decreasing = T), ]
names(PlotEcoCons)[1] <- "EVTYPE"

ggplot(PlotEcoCons[1:5, ], aes(EVTYPE, EcoConsDmg)) + geom_bar(stat = "identity") + ylab("Economic Damages (million dollars)") + 
  xlab("Event") + ggtitle("Top Five Events Causing Economic Damages Across the U.S")


```
  
From the diagram shows that flood damage causes the highest damage.  

### Conclusion  
The results show that, from year 1950 to November 2011, tornados are most harmful for population health and floods have the greatest economic losses.    

  
  